---
title: "Homework 7"
author: "Kent Roller"
date: "2024-05-19"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

3. Run cor.test() on the correlation between “area” and “perm” in the rock data set and  interpret the results. Note that you will have to use the  accessor to get at each of  the two variables (like this: rock$area). Make sure that you interpret both the confidence interval and the p-value that is generated by cor.test().

```{r}
rocks<-rock
cor.test(rock$area, rock$perm)
```

From the Pearson's test we obtain the CI of [-0.61,-0.13]. Here we can see that the CI does not encompass zero, meaning that our p-value should be very small. Looking at the p-value we verify this assumption, using an alpha of 0.05 for comparison, we can see that the p-value of 0.005254<0.05, as such, we reject the null hypothesis in favor of the alternative. We can also observe a moderate negative correlation of -0.40. \newpage


4. Create a copy of the bfCorTest() custom function presented in this chapter. Don’t forget to “source” it (meaning that you have to run the code that defines the function one  time to make R aware of it). Conduct a Bayesian analysis of the correlation between  “area” and “perm” in the rock data set.  
```{r}
library(BayesFactor)
bfCorTest <- function (x, y) {
  # Scale the variables
  zx <- scale(x)
  zy <- scale(y)
  
  # Create a data frame with the scaled variables
  zData <- data.frame(x = zx, rhoNot0 = zy)
  
  # Perform the Bayesian correlation test
  bfOut <- generalTestBF(x ~ rhoNot0, data = zData)
  
  # Extract the posterior samples
  mcmcOut <- posterior(bfOut, iterations = 10000)
  
  # Print the summary of the rhoNot0 column from the posterior samples
  print(summary(mcmcOut[, "rhoNot0"]))
  
  return(bfOut)
}


result <- bfCorTest(rock$area, rock$perm)
result
```

Using the output of the bfCorTest we can see that the HD interval generated, like the conventional CI does not encompass zero. This should result in a sizable rho ratio. When looking at the rhoNot0 output we can see that the resulting ratio is 8.1:1, which gives decent evidence to reject the null hypothesis in favor of the alternative. 
\newpage


8. Not unexpectedly, there is a data set in R that contains these data. The data set is  called UCBAdmissions and you can access the department mentioned above like  this: UCBAdmissions[, ,1]. Make sure you put two commas before the 1: this is a three  dimensional contingency table that we are subsetting down to two dimensions. Run  chisq.test() on this subset of the data set and make sense of the results. 
```{r}
admiss<-UCBAdmissions[, ,1]
admiss
chisq.test(admiss,correct=FALSE)
```

Observing the output from the Chi Squared test we can see that the X-squared value is fairly large, and that we run the test on 1 degree of freedom since this is a 2x2 contingency table. Looking at the p-value, we can see that it is very small, where 0.0000328<0.05. This means that we reject the null hypothesis of independence between gender and admission status. That is to say, the proportions of rejections between genders is not zero, or is not equal, where the proportion of rejections is higher among men than it is for women. 
\newpage 

9. Use contingencyTableBF() to conduct a Bayes factor analysis on the UCB admissions  data. Report and interpret the Bayes factor.
```{r}
ctBFadmiss<-contingencyTableBF(admiss, sampleType = "poisson", posterior=FALSE)
ctBFadmiss
```

From the output of the Bayes factor analysis we can see that we have very strong evidence in favor of the alternative hypothesis. Specifically, we obtain a ratio of 1111.64:1 that we should reject the null in favor of the alternative. Since this is much greater than the prerequisite 3:1 ratio, we can say that gender is not indepent of admissions decisions. 

























