---
title: "Intro to Data Science - HW 10 - Shiny"
output: html_document
runtime: shiny
---

##### Copyright Jeffrey Stanton, Jeffrey Saltz, Christopher Dunham, and Jasmina Tacheva


### Kent Roller 
### IST 687
### Due Date: 03/22/2024
### Submission Date: 03/22/2024

```{r}
# Enter your name here: Kent Roller 
```


### Attribution statement: (choose only one and delete the rest)

```{r}
# 1. I did this homework by myself, with help from the book and the professor.

```

At a high level, we want to create a scatter plot, where the user can select what is the x axis and the y axis of the scatter plot. We will use the same dataset as used in our first ggplot homework.

The size and color will be represented by the year attribute, and the user will be able to choose one of the following (for  the x and the y axis):
new_sp_m014
new_sp_f014
new_sp_m65
new_sp_f65

## Read in the same dataset we used for the ggplot homework

The file is: "https://intro-datascience.s3.us-east-2.amazonaws.com/who.csv" and store it in the *tb* dataframe
```{r}
library(tidyverse)
url<-"https://intro-datascience.s3.us-east-2.amazonaws.com/who.csv"
tb<-read_csv(url)
```
## Clean up the dataset, just like we did in the ggplot homework.

First, remove na's from iso2
```{r}
tb<-tb[!is.na(tb$iso2),]
```


Now create the dataframe 'tbCan', which is the iso2 for canada (CA)
```{r}
tb.can<-tb %>% filter(iso2 == "CA")
```

We will need the imputeTS package (only install if needed)
```{r}
#install.packages('imputeTS')
library(imputeTS)
```

Now we can use 'na_interpolation for new_sp_m014
```{r}
tb.can$new_sp_m014<-na_interpolation(tb.can$new_sp_m014)
tb.can$new_sp_m014[is.na(tb.can$new_sp_m014)]
```


## Create the User Interface of our shiny app

Define the sidebarPanel, which is two choices (use 'selectInput'), one for the x axis of the scatter plot, and the other is the y axis for the scatter plot. (make sure to library shiny).
```{r}
#install.packages("shiny")
library(shiny)
```



```{r}
#sidePanel <- sidebarPanel(
    #your code goes here
       # )

```

Create the mainPanel to show the scatter plot.

```{r}
#mainPanel <-  mainPanel(
    # your code goes here
       # )


# I have commented both of these chunks out because given how I understand how to create the apps this breaks it up in a way that I find confusing, so instead I will attempt to do the entire thing in the chunk below similar to how they were performed before 
```

# Define UI for application (called ui)

Use a 'fluidPage' using the sidebarLayout, and your sidePanel and mainPanel
```{r}
ui <- fluidPage(
  titlePanel("Scatterplot with Shiny"),
  sidebarLayout(
    sidebarPanel(
      selectInput("xVar", "Choose X variable:",
                  choices= names(tb.can)),
      selectInput("yVar", "Choose Y variable: ",
                  choices= names(tb.can))
    ),
    mainPanel(
      plotOutput("scatterPlot")
      
    )
  )
)

```

## Now let's define the server

Use ggplot to render a scatter plot, using the tbCan dataframe, the input for the x-axis and the input for the y-axis.

Store the results in the 'server' variable

```{r}

server <- function(input, output) {
  output$scatterPlot <- renderPlot({
    if(is.null(input$xVar) | is.null(input$yVar)) {
      return()
    }
    
    # Create scatterplot using ggplot
    ggplot(data = tb.can, aes_string(x = input$xVar, y = input$yVar)) +
      geom_point() + labs(x = input$xVar, y = input$yVar, title = paste("Scatterplot of", input$xVar, "vs.", input$yVar)) +
      theme_light()
  })
}

```

Now run the shiny App
```{r}
# Run the application 
shinyApp(ui = ui, server = server)
```
