## Intro to Data Science - HW 1
##### Copyright Jeffrey Stanton, Jeffrey Saltz, and Jasmina Tacheva
## IST 687
## Kent Roller 
## Homework 1 
## Due date:_/_/__
## Submission Date: 01/08/2024


```{r}
# Enter your name here: Kent Roller 
```

### Attribution statement: (choose only one and delete the rest)


```{r}
# 1. I did this homework by myself, with help from the book and the professor.
```

Define a variable:

```{r}
x <- 280
```

Define the following vectors, which represent the **population** (in thousands) and **number of colleges** in each of the five counties in Central New York (CNY) – **Cayuga**, **Cortland**, **Madison**, **Onondaga**, and **Oswego**, in this order:


```{r}
counties<- c("Cayuga","Cortland","Madison","Onondaga", "Oswego")
population <- c(80, 49, 73, 467, 122)
colleges <- c(2, 2, 3, 9, 2)
```

# Part 1: Calculating statistics using R  

A.	Show the number of observations in the **population** vector with the length() function: 


```{r}
length(population)
```

B.	Show the number of observations in the **colleges** vector with the length() function:


```{r}
length(colleges)
```

C.	Calculate the average CNY population using the mean() function:


```{r}
mean(population)
```

D.	Calculate the average number of colleges in CNY using the mean() function:


```{r}
mean(colleges)
```

E.	Calculate the total CNY population using the sum() function:


```{r}
sum(population)
```

F.	Calculate the total number of colleges in CNY using the sum() function:


```{r}
sum(colleges)
```

G.	Calculate the average CNY population again, this time using **the results from steps A & E**:


```{r}
sum(population)/length(population)
# or if you wanted them stored as variables then: 
popsum<-sum(population)
count<-length(population)
popsum/count
```

H.	Calculate the average number of colleges in CNY again, this time using **the results from steps B & F**:


```{r}
sum(colleges)/length(colleges)
# again, if you prefer it in variable format:
collegeTotal<-sum(colleges)
collegeCount<-length(colleges)
collegeTotal/collegeCount
```

# Part 2: Using the max/min and range functions in {r}

I.	How many colleges does the county with most colleges have? Hint: Use the max() function:


```{r}
max(colleges)
# can subset counties by colleges to get name of county as well:
counties[colleges==max(colleges)]

```

J.	What is the population of the least populous county in CNY? **Hint:** Use the min() function:


```{r}
min(population)
# as above we can subset counties by population parameters:
counties[population==min(population)]
```

K.	Display the populations of the least populous and most populous county in the dataset together. **Hint:** Use the range() function:


```{r}
range(population)
# Cant use the same method here because a range cant be applied over non multiples, also cant use & because the one value is not both the max and the min, however we can use the | argument to pass both to output as it looks for the min OR the max, and since we have both it outputs both. 

# I realize this wasn't explicitly asked for but I was afraid to turn it without it
counties[population==min(population) | population==max(population)]
```

# Part 3: Vector Math

L.	Create a new vector called **extraPop**, which is the current population of a county **+ 50** (each county has 50,000 more people):


```{r}
extraPop<-population + 50
```

M.	 Calculate the average of **extraPop**:


```{r}
mean(extraPop)
```

N.	In a variable called **bigCounties**, store all the population numbers from the original **population** vector which are **greater than 120** (using **subsetting** in R):


```{r}
bigCounties<-population[population > 120]
```

O.	Report the length of **bigCounties**:


```{r}
length(bigCounties)
```
