---
title: "Lab 4"
author: "Kent Roller"
date: "2023-03-26"
output:
  pdf_document: default
  html_document:
    df_print: paged
---
```{r}
library(SDaA)
data(coots)
attach(coots)
library(tidyverse)
```

4. One Stage Cluster Sampling 
Selecting a random sample of two clusters: 
```{r}
##clusterd<-1:100
##sample(clusterd,2)
## [1] 50 5
## Commenting out input after initial knit to prevent new numbers ##being generated each time because I knit multiple times per 
```

square #5: 8x5=40 
square #50: 10x5=50
N=100, n=2

Calculate the estimated mean, the standard error, and the estimated total: 

Cluster #5:\newline 
$M_i=40$\newline
$\bar{y}_i=40$\newline
$t_i=40$\newline
Cluster #50:\newline
$M_i=50$\newline
$\bar{y}_i=50$\newline
$t_i=50$\newline

$\widehat{\bar{y}}_r= \Sigma M_i\bar{y_i}/\Sigma M_i$\newline
=40(40)+50(50)/90\newline
=45.556\newline

$\bar{M}=45$\newline
$M^2_i(\bar{y}-\widehat{\bar{y}}_r)^2$\newline
Cluster 5= 40^2 (40-45.66)^2 = 51257\newline    
Cluster 50= 50^2 (50-45.66)^2 = 47089\newline
$\Sigma M^2_i(\bar{y}-\widehat{\bar{y}}_r)^2$=98346\newline
SE=sqrt(1-(2/100))(1/((2)(45)^2)(98346/1)$\approx 4.878$\newline

N/n=100/2=50 \newline
$\widehat{t}_{unb}$\newline
=50(90)=4500\newline

Estimated Mean= 45.66\newline
Standard Error= 4.878\newline
Estimated Total= 4500\newpage

All equations used reflect the fact that the clusters are of uneven sizes, 5.16, 5.17, and 5.12 were used (in order)\newline

Evidently I am supposed to have an n=20 not n=2 according to one of the problems below. So the above is for n=2, I will add the work for n=20 below. 
```{r}
clusterd<-1:10
sample(clusterd,2)
#first output was 3 and 4
```

N=100, n=20
clusters 3 and 4

Cluster 3:\newline
Area 31:60\newline
Area 32:18\newline
Area 33:3\newline
Area 34:60\newline
Area 35:35\newline
Area 36:36\newline
Area 37:6\newline
Area 38:12\newline
Area 39:30\newline
Area 40:20\newline

Cluster 4:\newline 
Area 41:45\newline
Area 42:5\newline
Area 43:12\newline
Area 44:40\newline
Area 45:2\newline
Area 46:21\newline
Area 47:8\newline
Area 48:54\newline
Area 49:27\newline
Area 50:50\newline 

Cluster #5:\newline 
$M_i=40$\newline
$\bar{y}_i=40$

Calculate the estimated mean, the standard error, and the estimated total\newline

Estimated total:\newline
$\widehat{\bar{y}}_r= \Sigma M_i\bar{y_i}/\Sigma M_i$\newline
```{r}
area.c<-c(60,18,3,60,35,36,6,12,30,20,45,5,12,40,2,21,8,54,27,50)
M.i<-20
bar.y<-sum(area.c)/M.i
hat.y<-20*(bar.y)/M.i
hat.y
bar.y
N.n<-100/20
t.unb<-N.n*(sum(area.c))
t.unb


```


```{r}
sqrt((1-(20/100))*(1/((20)*(33.9^2)))*(1690373/(20-1)))
```
Standard Error=1.759725

$\widehat{t}_{unb}=$


$\widehat{\bar{y}}_r$=27.2\newline
SE=1.759725\newline
Total Estimate=2720\newline

From lab 2 had the following information:\newline
$\bar{y}=34.1$\newline
Total Area=2720\newline
The estimates for the average value are the same as the values found in lab 2.  \newpage 



6. The point estimator for totals and proportions: \newline 
What other information would you need in order to estimate the total yield of the whole population of 150 farms?\newline
We would want an measure of the Mean Square Within and Between to make sure the difference between the measurements obtained wasn't so great that we would be unable to make a determination from it. We would also want to know the ICC to give us an idea of the homogeneity within the clusters. This would also let us know if the sample size we have selected will yield a high enough precision to be useful, or whether we should consider a higher sample size or a different testing method altogether such as SRS. Additionally, is an acre a large enough size for each measurement to be able to accurately produce a usable estimate? \newline

How would you obtain from equation 5.15 a point estimator for a population proportion when the variable of interest is categorical (yes/no) type?\newline
For non categorical data (quantitative) we can use unbiased estimators to calculate a point estimator for the population proportion. However, when we are unable to use unbiased estimators, such as in cases like this question where the variable of interest is categorical we can use Ratio Estimation to estimate what proportion of responses were yes or no.\newpage


8. Example 5.6 from the book\newline
```{r}
M<-c(20,26,24,34,26,28,19,32,17,21,26,26)
ybar<-c(61.5,64.2,58.4,58,58,64.9,55.2,72.1,58.2,66.6,62.3,67.2)

```
Write the necessary R code to estimate the mean score in the population (equation 5.15)\newline
$\widehat{\bar{y}}_r= \Sigma M_i\bar{y_i}/\Sigma M_i$\newline
```{r}
sum(ybar*M)/sum(M)
#[1] 62.57191
```

Write the necessary R code to calculate the standard error (equation 5.17) \newline
```{r}
sqrt((1-(12/187))*(1/((12)*(24.92^2)))*(194827/(12-1)))
#[1] 1.491378
```

Code provides same answer as the book 
\newpage

9. Two Stage Clustering Sample\newpage

10. \newpage

11. The egg example\newline

The example on pages 186-188 is that of the mean volume of eggs. The clusters are clutches of eggs and the individuals or secondary sampling units are the eggs. The variable of interest is volume. 
Assume you are given the following information:\newline
$\#$ of eggs in each clutch\newline
Mean volume of the eggs in the sample for each clutch\newline 
Standard deviation in each sample\newline

What R commands would allow you to do the calculations on page 188 
```{r}
Mi<-csize
y.bar<-tapply(coots$volume,coots$clutch,mean)
n.csize<-csize[seq(1,length(csize),2)]
t.est<-n.csize*y.bar
t.tot<-sum(t.est)
Mi.s<-sum(n.csize)

```
To find $\widehat{\bar{y_r}}$
```{r}
y.est<-t.tot/Mi.s
y.est
```
$\widehat{\bar{y_r}}=2.4906$\newline 

To find $s^2_r$:
```{r}
#s.i.q[i]<-(coots$volume[2*i-1]-y.est[i])^2+(coots$volume[2*i]-y.bar[i])^2
```









