---
title: "Chapter 12 Lab"
author: "Kent Roller"
date: "2023-04-29"
output:
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Need to find the total proportion of the population that has diabetes. 
To do this will use : \newline 
$\widehat{t}^2_{str}=N/n\Sigma (n_h/m_h)r_h$\newline
1st\newline
```{r}
a<-(241/96)*(86)
b<-(112/45)*(17)
c<-(174/35)*(29)
d<-(472/47)*(8)
N.n<-100000/1000
est<-N.n*sum(a+b+c+d)
est
```


So based on this estimate the total number of the population that has diabetes is 48271.88.

Using Raos estimator for variance: 
```{r}
N<-100000
n<-1000
V<-N*(N-1)*0.00089313+(N^2/n-1)*(1-n/n)*0.992310
se<-sqrt(V)
se
lower<-est-qnorm(0.975)*se
upper<-est+qnorm(0.975)*se

upper
lower
```

For a 95% confidence interval we have [42414.5,54129.26]. Here we can see that the total estimate falls within our confidence interval. What this means is that our confidence interval of between 42415 and 54129 members of the population will capture the true mean of 48271 members of the population 95% of the time. 
