---
title: "STAT 4037 Lab1 R"
date: "2023-01-27"
output:
  pdf_document: default
  pdf: default
---
## Name: Kent Roller
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Reading in the data set using file.choose and read.csv because I kept getting errors about empty columns using read table and when I attempted to pass the fill=TRUE argument into the command it botched the data set.  

```{r}
file.choose()

drugs<-read.csv("C:\\Users\\super\\Downloads\\drugsurv.csv", header=TRUE)

head(drugs)

attach(drugs)

names(drugs)
```

## Table and Graph for Gender

```{r}
table(Gender)
counts<-table(Gender)
barplot(counts)
pie(counts)
```

## Creating a Histogram for age

```{r}
hist(Age)
hist(Age, col="red")
```

## Recoding age in three groups:
## If there are only two classes to use:
```{r}
agenew<-ifelse(Age<50, "young", "old")
table(agenew)
```

## If there are THREE groups
```{r}
agegroup3<-ifelse(Age<40, "young", ifelse(Age>40 & Age<60, "middle aged", "old"))
table(agegroup3)
```


## Creating a new Quantitative Variable 
```{r}
agenow<-Age+15
```


## Cross Tabulation of gender and marijuana
```{r}
counts<-table(Gender,Mariguana)
barplot(counts, beside=TRUE, legend=rownames(counts))
barplot(counts, beside=FALSE, legend=rownames(counts))
```

## Creating a subset formed just by the ever smokers 
```{r}
smokersonly<-drugs[Ever.Smoker=="YES",]
head(smokersonly)
hist(smokersonly$Age)
```

## Calculating correlation between age and years of education 
```{r}
cor(Age,Education)
plot(Age,Education)
plot(Age, Education, xlab="Age", ylab="Years of Education")
```

## More things to do with R
```{r}
mean(Age)
var(Age)
summary(Age)
sd(Age)
tapply(Age, Gender, summary)
boxplot(Age)
boxplot(Age~Gender, horizontal=TRUE, xlab="Age")
mytable<-table(Gender,Smoking.Status)
mytable
margin.table(mytable,1)
margin.table(mytable,2)
prop.table(mytable,1)
prop.table(mytable,2)
```



